Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	22:43:04	(Tue 6.Oct)

MVDR seeks to minimize the output power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}. For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}. Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

\begin{figure}[tbp]%
\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected by the 32 element HISAS1030 sonar attached to the HUGIN AUV.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image. The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale. The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight. The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
\end{figure*}

@@

If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align}
E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. \todo{Might want to be careful with this when we steer the windows, or drop it altogether}
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{eq:kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window. A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as 
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
}.\label{eq:steering_vector}
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

There is an upper bound to the amount of steering we should apply to a window. Since each window's response is constrained to a unit gain in the look direction, the steered window will have increased white noise gain and deteriorated signal-to-noise ratio as shown in \Fig{windows_steering}. To limit this we placing an upper bound to the steering, realizing that  a wide window will be allowed more steering than a narrow one.

We call the upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. The displayed images are all upinterpolated to the same size, but the detail level varies. 
% 
% \todo{Need $M$ in the image sector. Why???}

@@

Two different scenes will be presented. One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}. It measures 68\;m by 9\;m and fills most of the highlighted sector. This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime. The other scene is of an iron cross of 1\,m length in both directions, connected to a an anchor with a 13\,cm diameter (\Fig{cross}). This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr, Maritime Technology and Research (WTD 71). Both scenes is imaged in sectorscan mode.

In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading. Both MVDR and LCA was run with $K=1$ temporal averaging. This is a fairly aggressive yet stable set of parameters.\todo{suitable ref?}

The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize  that these can be mimicked by Kaiser windows. In section \ref{sec:results_window_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be . In section \ref{sec:results_database_size} we discuss how many window variations is needed for LCA to perform well.

@@

Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with with responses similar to the ones MVDR computes. We study this in \Fig{mvdr_selected_windows}, where we present the typical responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.

Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones. The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor. We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which is required to allow MVDR to run in near snapshot mode in an active system. The peak sidelobe levels of the MVDR windows are mostly between -13\,dB and -43\,dB. This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.

>>>>

MVDR seeks to minimize the output power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

\begin{figure}[tbp]%
\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image. The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale. The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight. The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
\end{figure*}

@@

\todo{Drop this paragraph altogether?}If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align}
E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. 
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{eq:kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time\todo{Should perhaps say space-bandwidth product?} and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced\todo{better word?} sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window. A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as 
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
}.\label{eq:steering_vector}
\end{align}
%
Here $d$\todo{$d$ is reused as element spacing, find an alternative} is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

Each window's response is constrained to unit gain in the look direction. Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}. To limit this we place an upper bound to the steering, chosen such that wide windows is allowed to be steered more than narrow ones. We call this upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images are all bilinearly upinterpolated to the same size. 
% 
% \todo{Need $M$ in the image sector. Why???}

@@

Two different scenes will be studied. One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}. It measures 68\;m by 9\;m and fills most of the highlighted sector. This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime. The other scene is of an 1\,m by 1\,m iron cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}). This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71). Both scenes are imaged in sectorscan mode.

In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading\todo{also used for element distance}. Both MVDR and LCA was run with $K=1$ temporal averaging. This is a fairly aggressive yet stable set of parameters.\todo{suitable ref?}

The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows. In section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$. In section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.

@@

Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with with responses similar to the ones that MVDR computes. We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.

Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones. The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor. We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate in near snapshot mode in an active system. The peak sidelobe levels of the MVDR windows are mostly between \minus{}13\,dB and \minus{}43\,dB. This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.
